New cardinality estimation algorithms for HyperLogLog sketches

### 2016-09-26 paper/paper.tex
fixed some minor issues, reformatted some equations

The logarithm of the joint probability mass function can be written using Iverson bracket notation ($\left[\text{true}\right]:=1$, $\left[\text{false}\right]:=0$) as
\begin{align}
\label{equ:joint_log_pmf_single_register}
\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
=&
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
\left[\symRegVal_1\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
\left[\symRegVal_2\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1,\symRegVal_2\right)}}
\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
\left[1\leq\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
\left[\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
\left[1\leq\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
\left[\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\!\left(
1
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+
e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
\right)
\left[1\leq\symRegVal_1=\symRegVal_2\right].
\end{align}

@@

\begin{align}

@@

@@

\nonumber\\

@@

\nonumber\\
\nonumber\\
\nonumber\\
\symCountVariate^{=}_{\symRegVal},
\end{align}

@@

The transformed log-likelihood function can be written as
%\begin{multline}
%\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
%\log \symLikelihood(
%\symNumReg e^{\symPhi_\symSetASuffix},
%\symNumReg e^{\symPhi_\symSetBSuffix},
%\symNumReg e^{\symPhi_\symSetXSuffix}
%\vert
%\boldsymbol{\symRegValVariate}_1,
%\boldsymbol{\symRegValVariate}_2
%)=
%\\
%\begin{aligned}
%&-
%\sum_{\symRegVal=0}^{\symRegRange}
%\left(
%  \symCountVariate^{<}_{1\symRegVal}+
%  \symCountVariate^{=}_{\symRegVal}+
%  \symCountVariate^{>}_{1\symRegVal}
%\right)
%\frac{e^{\symPhi_\symSetASuffix}}{2^{\symRegVal}}
%+
%\left(
%  \symCountVariate^{<}_{2\symRegVal}+
%  \symCountVariate^{=}_{\symRegVal}+
%  \symCountVariate^{>}_{2\symRegVal}
%\right)
%\frac{e^{\symPhi_\symSetBSuffix}}{2^{\symRegVal}}
%+
%\left(
%  \symCountVariate^{<}_{1\symRegVal}+
%  \symCountVariate^{=}_{\symRegVal}+
%  \symCountVariate^{<}_{2\symRegVal}
%\right)
%\frac{e^{\symPhi_\symSetXSuffix}}{2^{\symRegVal}}
%\\
%&+
%\sum_{\symRegVal=1}^{\symRegRange}
%\symCountVariate^{<}_{1\symRegVal}
%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
%+e^{\symPhi_\symSetXSuffix}
%}{2^{\symRegVal}}}\right)
%+
%\symCountVariate^{<}_{2\symRegVal}
%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
%+
%e^{\symPhi_\symSetXSuffix}
%}{2^{\symRegVal}}}\right)
%\\
%&+
%\sum_{\symRegVal=1}^{\symRegRange+1}
%\symCountVariate^{>}_{1\symRegVal}
%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetASuffix}
%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
%+
%\symCountVariate^{>}_{2\symRegVal}
%\log\!\left(1-e^{-\frac{e^{\symPhi_\symSetBSuffix}
%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}\right)
%\\
%&+
%\sum_{\symRegVal=1}^{\symRegRange+1}
%\symCountVariate^{=}_{\symRegVal}
%\log\!\left(
%1
%-e^{-\frac{e^{\symPhi_\symSetASuffix}
%+e^{\symPhi_\symSetXSuffix}
%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
%-
%e^{-\frac{e^{\symPhi_\symSetBSuffix}
%+e^{\symPhi_\symSetXSuffix}
%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
%+
%e^{-\frac{e^{\symPhi_\symSetASuffix}
%+e^{\symPhi_\symSetBSuffix}
%+e^{\symPhi_\symSetXSuffix}
%}{2^{\min\left(\symRegVal,\symRegRange\right)}}}
%\right).
%\end{aligned}
%\end{multline}

\begin{multline}
\label{equ:max_likelihood_phi_func}
\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
\log \symLikelihood(
\symNumReg e^{\symPhi_\symSetASuffix},
\symNumReg e^{\symPhi_\symSetBSuffix},
\symNumReg e^{\symPhi_\symSetXSuffix}
\vert
\boldsymbol{\symRegValVariate}_1,
\boldsymbol{\symRegValVariate}_2
)=
\\
\begin{aligned}
&+
\sum_{\symRegVal=1}^{\symRegRange}
\symCountVariate^{<}_{1\symRegVal}
\log\!\left(
\symZ_{\symSetXSuffix\symRegVal}
+
\symY_{\symSetXSuffix\symRegVal}
\symZ_{\symSetASuffix\symRegVal}
\right)
+
\symCountVariate^{<}_{2\symRegVal}
\log\!\left(
\symZ_{\symSetXSuffix\symRegVal}
+
\symY_{\symSetXSuffix\symRegVal}
\symZ_{\symSetBSuffix\symRegVal}
\right)
\\
&+
\sum_{\symRegVal=1}^{\symRegRange}
\symCountVariate^{>}_{1\symRegVal}
\log\!\left(
\symZ_{\symSetASuffix\symRegVal}
\right)
+
\symCountVariate^{>}_{2\symRegVal}
\log\!\left(
\symZ_{\symSetBSuffix\symRegVal}
\right)
+
\symCountVariate^{=}_{\symRegVal}
\log\!\left(
\symZ_{\symSetXSuffix\symRegVal}
+
\symY_{\symSetXSuffix\symRegVal}
\symZ_{\symSetASuffix\symRegVal}
\symZ_{\symSetBSuffix\symRegVal}
\right)
\\
&+
\symCountVariate^{>}_{1,\symRegRange+1}
\log\!\left(
\symZ_{\symSetASuffix\symRegRange}
\right)
+
\symCountVariate^{>}_{2,\symRegRange+1}
\log\!\left(
\symZ_{\symSetBSuffix\symRegRange}
\right)
+
\symCountVariate^{=}_{\symRegRange+1}
\log\!\left(
\symZ_{\symSetXSuffix\symRegRange}
+
\symY_{\symSetXSuffix\symRegRange}
\symZ_{\symSetASuffix\symRegRange}
\symZ_{\symSetBSuffix\symRegRange}
\right)
\\
&-
\sum_{\symRegVal=0}^{\symRegRange}
\left(
  \symCountVariate^{<}_{1\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{>}_{1\symRegVal}
\right)
\symX_{\symSetASuffix\symRegVal}
+
\left(
  \symCountVariate^{<}_{2\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{>}_{2\symRegVal}
\right)
\symX_{\symSetBSuffix\symRegVal}
+
\left(
  \symCountVariate^{<}_{1\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{<}_{2\symRegVal}
\right)
\symX_{\symSetXSuffix\symRegVal}
.
\end{aligned}
\end{multline}
Here we introduced the following expressions for simplification:
\begin{equation}
\label{equ:definition_xyz}
\symX_{\ast\symRegVal} :=\frac{e^{\symPhi_\ast}}{2^{\symRegVal}},\quad
\symY_{\ast\symRegVal} := e^{-\symX_{\ast\symRegVal}},\quad
\symZ_{\ast\symRegVal} := 1 - \symY_{\ast\symRegVal}.
\end{equation}

@@

The calculation of \eqref{equ:max_likelihood_phi_func} and its derivatives requires some care when calculating $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$. Since $\symX_{\ast\symRegVal}$ is nonnegative, the value range of both is $[0,1]$. In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\texttt{expm1}(\symX):=e^\symX-1$ that is available in most programming languages. If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} = 
-\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}. In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and only a single exponential function evaluation is needed.

@@

\subsection{Results}
We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm to solve this optimization. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.

TODO see \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x}

>>>>

The logarithm of the joint probability mass function can be written using Iverson bracket notation ($\left[\text{true}\right]:=1$, $\left[\text{false}\right]:=0$) as
\begin{multline}
\label{equ:joint_log_pmf_single_register}
\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
=\\
\begin{aligned}
&
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
\left[1\leq\symRegVal_1<\symRegVal_2\right]
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
\left[\symRegVal_2<\symRegVal_1\right]
\\
+
&
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
\left[1\leq\symRegVal_2<\symRegVal_1\right]
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
\left[\symRegVal_1<\symRegVal_2\right]
\\
+
&
\log\!\left(
1
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+
e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
\right)
\left[1\leq\symRegVal_1=\symRegVal_2\right]
\\
-
&
\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
\left[\symRegVal_1\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
\left[\symRegVal_2\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1,\symRegVal_2\right)}}
\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
\end{aligned}
\end{multline}

@@

@@

\symCountVariate^{=}_{\symRegVal}

@@

@@

}{2^{\symRegVal}},

@@

The transformed log-likelihood function can be written as
\begin{multline}
\label{equ:max_likelihood_phi_func}
\symFunc(\symPhi_\symSetASuffix,\symPhi_\symSetBSuffix,\symPhi_\symSetXSuffix):=
\log \symLikelihood(
\symNumReg e^{\symPhi_\symSetASuffix},
\symNumReg e^{\symPhi_\symSetBSuffix},
\symNumReg e^{\symPhi_\symSetXSuffix}
\vert
\boldsymbol{\symRegValVariate}_1,
\boldsymbol{\symRegValVariate}_2
)=
\\
\begin{aligned}
&+
\sum_{\symRegVal=1}^{\symRegRange}
\symCountVariate^{<}_{1\symRegVal}
\log\!\left(
\symZ_{\symSetXSuffix\symRegVal}
+
\symY_{\symSetXSuffix\symRegVal}
\symZ_{\symSetASuffix\symRegVal}
\right)
+
\symCountVariate^{<}_{2\symRegVal}
\log\!\left(
\symZ_{\symSetXSuffix\symRegVal}
+
\symY_{\symSetXSuffix\symRegVal}
\symZ_{\symSetBSuffix\symRegVal}
\right)
\\
&+
\sum_{\symRegVal=1}^{\symRegRange}
\symCountVariate^{>}_{1\symRegVal}
\log\!\left(
\symZ_{\symSetASuffix\symRegVal}
\right)
+
\symCountVariate^{>}_{2\symRegVal}
\log\!\left(
\symZ_{\symSetBSuffix\symRegVal}
\right)
+
\symCountVariate^{=}_{\symRegVal}
\log\!\left(
\symZ_{\symSetXSuffix\symRegVal}
+
\symY_{\symSetXSuffix\symRegVal}
\symZ_{\symSetASuffix\symRegVal}
\symZ_{\symSetBSuffix\symRegVal}
\right)
\\
&+
\symCountVariate^{>}_{1,\symRegRange+1}
\log\!\left(
\symZ_{\symSetASuffix\symRegRange}
\right)
+
\symCountVariate^{>}_{2,\symRegRange+1}
\log\!\left(
\symZ_{\symSetBSuffix\symRegRange}
\right)
+
\symCountVariate^{=}_{\symRegRange+1}
\log\!\left(
\symZ_{\symSetXSuffix\symRegRange}
+
\symY_{\symSetXSuffix\symRegRange}
\symZ_{\symSetASuffix\symRegRange}
\symZ_{\symSetBSuffix\symRegRange}
\right)
\\
&-
\sum_{\symRegVal=0}^{\symRegRange}
\left(
  \symCountVariate^{<}_{1\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{>}_{1\symRegVal}
\right)
\symX_{\symSetASuffix\symRegVal}
+
\left(
  \symCountVariate^{<}_{2\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{>}_{2\symRegVal}
\right)
\symX_{\symSetBSuffix\symRegVal}
+
\left(
  \symCountVariate^{<}_{1\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{<}_{2\symRegVal}
\right)
\symX_{\symSetXSuffix\symRegVal}
.
\end{aligned}
\end{multline}
Here we introduced the following expressions for simplification:
\begin{equation}
\label{equ:definition_xyz}
\symX_{\ast\symRegVal} :=\frac{e^{\symPhi_\ast}}{2^{\symRegVal}},\quad
\symY_{\ast\symRegVal} := e^{-\symX_{\ast\symRegVal}},\quad
\symZ_{\ast\symRegVal} := 1 - \symY_{\ast\symRegVal}.
\end{equation}

@@

The calculation of \eqref{equ:max_likelihood_phi_func} and its derivatives requires some care when calculating $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$. Since $\symX_{\ast\symRegVal}$ is nonnegative, the value range of both is $[0,1]$. In order to reduce the numerical error of $\symZ_{\ast\symRegVal}$ for small $\symX_{\ast\symRegVal}$, it is essential to use the function $\texttt{expm1}(\symX):=e^\symX-1$ that is available in most programming languages. If $\symX_{\ast\symRegVal}$ is smaller than $\log(2)$, we calculate $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$ via $\symZ_{\ast\symRegVal} = 
-\texttt{expm1}(-\symX_{\ast\symRegVal})$ and $\symY_{\ast\symRegVal} = 1 - \symZ_{\ast\symRegVal}$ and not as defined in \eqref{equ:definition_xyz}. In this way the numerical error of both, $\symY_{\ast\symRegVal}$ and $\symZ_{\ast\symRegVal}$, is minimized and only a single exponential function needs to be evaluated.

@@

\subsection{Results}
We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm to solve this optimization problem. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.

TODO see results \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x}
